Tracking $p$-adic precision

### 2014-05-16 padicprec.tex
More changes suggested by referee

On the other hand, one can prove that the SOMOS 4 sequence 
exhibits the \emph{Laurent phenomenon} \cite{fomin-zelevinsky:02a}: for all integer 
$n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
From the latter formula, it follows directly that if $u_0$, $u_1$,
$u_2$ and $u_3$ are known up to precision $O(p^N)$ then all $u_n$'s
are also known with the same precision. Thus, the term $v_n$ that 
appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
precision but some numerical instability related to the algorithm used to compute $u_n$.

@@

Let us first recall that a square matrix $M$
admits a LU factorization if it can be written as a product $LU$ where 
$L$ and $U$ are lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems, and is
discussed further in Appendix \ref{ssec:matrices}.
When computing the entries of $L$ and $U$ from a $d \times d$ matrix
over $\Zp$ with entries of precision $O(p^N)$, one has a choice of algorithms:
\begin{itemize}
\item using usual Gaussian elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulae yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$ \cite{caruso:12a}.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. On the other hand, the second algorithm
is less efficient than the first one because evaluating Cramer-type
formulae requires many computations. 
%In \cite{caruso:12a}, the author designs a new algorithm which combines 
%both advantages: efficiency and stability.
%The precision loss inherent in an algorithm is an important factor in 
%analyzing its performance.
%However, since this precision loss depends heavily on the details of the 
%problem and the algorithm, we will not focus on such questions in this 
%paper.

@@

Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
We assume that the induced metric is an ultrametric (\emph{i.e.}
$\lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert)$)
and that $K$ is complete with respect to it.
For example, we may take $K = \Qp$ with the
$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
that $K$ contains a dense subring $R \subset K$
consisting of elements that can be represented with a finite amount of space.
For $K = \Qp$ we may choose $R = \Z[\frac1p]$ or $R = \Q$;
for $K = \Fp\lp t \rp$ we may choose $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$.

If $E$ is a $K$-vector space, possibly of infinite dimension, then an
\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A \emph{$K$-Banach space} is a complete normed $K$-vector space.  Note that any finite-dimensional
normed $K$-vector space is automatically complete and all norms over such a space are equivalent.
A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
We underline that any lattice $H$ in $E$ is also closed since its 
complement is the union of all cosets $a + H$ (with $a \not\in H$) which 
are all open.
For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
$$B_E(r) = \{ x \in E : \Vert x \Vert \le r \}, 
\quad B^-_E(r) = \{ x \in E : \Vert x \Vert < r\}.$$
Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.

Suppose $E$ is a $K$-Banach space and $I$ a set.
A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
every element $x \in E$ can be written $x = \sum_{i \in I} \alpha_i x_i$
for scalars $\alpha_i \in K$ with $\alpha_i \to 0$, and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.

@@

For the rest of this section, we assume that $K$ is algebraically closed. As in Definition 
\ref{def:locanal}, we consider two $K$-Banach spaces $E$ and $F$ and a 
family of continuous $n$-linear maps $L_n : E^n \to F$. For $n \ge 1$ 
and $h \in E$, we set $f_n(h) = L_n(h, \ldots, h)$ and
$$\Vert f_n \Vert = \sup_{h \in B_E(1)} \Vert f_n(h) \Vert.$$
When the series $\sum_n f_n(h)$ converges, we denote by $f(h)$ its sum;
we shall write $f = \sum_{n \geq 0} f_n$. We assume that $f$ is defined
in a neighborhood of $0$. Under this assumption, the datum of $f$ 
uniquely determines the $f_n$'s (a consequence of Proposition 
\ref{prop:lambdaNP} below).
To such a series $f$, we attach the function $\Lambda(f) : \R \to \R 
\cup \{+\infty\}$ defined by:
$$\begin{array}{rcll}
\Lambda(f)(v) & = & 
\log \big( \sup_{h \in B^-_E(e^v)} \Vert f(h) \Vert \big)
& \text{if } f \text{ is defined on } B^-_E(e^v) \\
& = & +\infty & \text{otherwise}
\end{array}$$
The following lemma is easy and left to the reader.

\begin{lem}
\label{lem:opLambda}
Let $f = \sum_{n \geq 0} f_n$ and $g = \sum_{n \geq 0} g_n$ be two 
series as above. Then:
\begin{align*}
\Lambda(f+g) &\leq \max (\Lambda(f),\Lambda(g)) \\
\Lambda(f \times g) &\leq \Lambda(f)+\Lambda(g) \\
\Lambda(f \circ g) &\leq \Lambda(f) \circ \Lambda(g)
\end{align*}
\end{lem}

@@

Recall from \S \ref{ssec:stepbystep} that a SOMOS 4 sequence is a 
four-term inductive sequence defined by $u_n = \frac{u_{n-3} u_{n-1} 
+ u_{n-2}^2}{u_n}$ exhibiting the Laurent phenomenon.
We will focus on SOMOS sequences with values in $\Q_p$, and 
assume for simplicity that $u_0, u_1, u_2, u_3 \in \Z_p^\times$.
By the Laurent phenomenon, $u_n \in \Z_p$ for all $n$, and if $u_0, u_1, u_2, u_3$ 
are known with finite precision $O(p^N)$ then all $u_n$ are 
known with the same absolute precision. Algorithm \ref{algo:SOMOS} presented on 
page \pageref{algo:SOMOS} performs this computation.

@@

Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We may
consider the map sending a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{vaccon:14b}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from \cite{vaccon:14b}*{Thm. 1.1} that this 
map is continuous at $(f_1, \ldots, f_s)$ provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is weakly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{vaccon:14b} for more detail).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is an $(s \times t)$ matrix with coefficients in $K[\XX]$.
Following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating \eqref{eq:prodgrobner}, we get
$$(d g_1, \ldots, d g_t) =
(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
Finally, since $(g_1, \ldots, g_t)$ is a \emph{reduced} Gr\"obner
basis, the above equality implies that $d g_i$ is the remainder in
the division of the $i$-th entry of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.

>>>>

On the other hand, one can prove that the SOMOS 4 sequence 
exhibits the \emph{Laurent phenomenon} \cite{fomin-zelevinsky:02a}: for all integer 
$n$, there exists a polynomial $P_n$ in $\Z[X^{\pm 1}, Y^{\pm 1}, Z^{\pm 
1}, T^{\pm 1}]$ such that $u_n = P_n(u_0, u_1, u_2, u_3)$.
From the latter formula and our assumption that $u_0$, $u_1$, 
$u_2$ and $u_3$ are units known up to precision $O(p^N)$, it follows that all $u_n$'s
are known with the same precision. Thus, the term $v_n$ that 
appears in \eqref{eq:SOMOS} does not reflect an intrinsic loss of
precision but some numerical instability related to the algorithm used to compute $u_n$.

@@

Let us first recall that a square matrix $M$
admits a LU factorization if it can be written as a product $LU$ where 
$L$ and $U$ are lower triangular and upper triangular respectively. 
The computation of a LU factorization appears as an important tool to 
tackle many classical questions about matrices or linear systems, and is
discussed further in Appendix \ref{ssec:matrices}.
When computing the entries of $L$ and $U$ from a $d \times d$ matrix
over $\Zp$ with entries of precision $O(p^N)$, one has a choice of algorithms:
\begin{itemize}
\item using usual Gaussian elimination and tracking $p$-adic precision 
step-by-step, the smallest precision on an entry of $L(M)$ is about
$O(p^{N - \frac{2d}{p-1}})$ on average;
\item computing $L(M)$ by evaluating Cramer-type formulae yields a
result whose every entry is known up to precision $O(p^{N - 2 \log_p 
d})$ \cite{caruso:12a}.
\end{itemize}
If $d$ is large compared to $p$, the second precision is much more
accurate than the first one. On the other hand, the Cramer-type formulae
in the second algorithm yield a substantially longer running time than the first.
%In \cite{caruso:12a}, the author designs a new algorithm which combines 
%both advantages: efficiency and stability.
%The precision loss inherent in an algorithm is an important factor in 
%analyzing its performance.
%However, since this precision loss depends heavily on the details of the 
%problem and the algorithm, we will not focus on such questions in this 
%paper.

@@

Let $K$ be a field with absolute value $|\cdot| : K \to \R_{\ge 0}$.
We assume that the induced metric is an ultrametric (\emph{i.e.}
$\lvert x + y \rvert \leq \max(\lvert x \rvert, \lvert y \rvert)$)
and that $K$ is complete with respect to it.
For example, we may take $K = \Qp$ with the
$p$-adic absolute value or $K = k\lp t \rp$ with the $t$-adic absolute value.
Write $\OK$ for the ring $\{x \in K : \lvert x \rvert \le 1\}$ and assume
that $K$ contains a computable dense subring $R \subset K$ \cite{rabin:60a}.
This assumption holds for $K = \Qp$ and $K=\Fp\lp t \rp$ by setting $R = \Z[\frac1p]$ or $R = \Q$
in the case of $\Qp$ and $R = \Fp[t, t^{-1}]$ or $R = \Fp(t)$ in the case of $\Fp\lp t \rp$.

If $E$ is a $K$-vector space, possibly of infinite dimension, then an
\emph{ultrametric norm} on $E$ is a map $\Vert\cdot\Vert : E \to \R^+$ satisfying:
\begin{enumerate}[(i)]
\item $\Vert x\Vert = 0$ if and only if $x = 0$;
\item $\Vert \lambda x\Vert = |\lambda| \cdot \Vert x\Vert$;
\item $\Vert x+y\Vert \leq \max(\Vert x\Vert, \Vert y\Vert)$.
\end{enumerate}
A \emph{$K$-Banach space} is a complete normed $K$-vector space.  Note that any finite-dimensional
normed $K$-vector space is automatically complete and all norms over such a space are equivalent.
A \emph{lattice} in a $K$-Banach space $E$ is an open bounded sub-$\OK$-module of $E$.
We underline that any lattice $H$ in $E$ is also closed since its 
complement is the union of all cosets $a + H$ (with $a \not\in H$) which 
are all open.
For a $K$-Banach space $E$ and $r \in \R_{\ge 0}$, write
$$B_E(r) = \{ x \in E : \Vert x \Vert \le r \}, 
\quad B^-_E(r) = \{ x \in E : \Vert x \Vert < r\}.$$
Note that $B_E(r)$ and $B^-_E(r)$ are both lattices.  We will also set $B_E(\infty) = B^-_E(\infty) = E$.

Suppose $E$ is a $K$-Banach space and $I$ a set.
A family $(x_i)_{i \in I} \subset E$ is a \emph{Banach basis} for $E$ if
every element $x \in E$ can be written $x = \sum_{i \in I} \alpha_i x_i$
for scalars $\alpha_i \in K$ with $\alpha_i \to 0$ (according to the filter of cofinite subsets), and $\Vert x \Vert = \sup_{i \in I} \lvert \alpha_i \rvert$.
Note that if $E$ is finite dimensional then the condition $\alpha_i \to 0$ is vacuous.

@@

For the rest of this section, we assume that $K$ is algebraically closed. As in Definition 
\ref{def:locanal}, we consider two $K$-Banach spaces $E$ and $F$ and a 
family of continuous $n$-linear maps $L_n : E^n \to F$. For $n \ge 1$ 
and $h \in E$, we set $f_n(h) = L_n(h, \ldots, h)$ and
$$\Vert f_n \Vert = \sup_{h \in B_E(1)} \Vert f_n(h) \Vert.$$
When the series $\sum_n f_n(h)$ converges, we denote by $f(h)$ its sum;
we shall write $f = \sum_{n \geq 0} f_n$. We assume that $f$ is defined
in a neighborhood of $0$. Under this assumption, the datum of $f$ 
uniquely determines the $f_n$'s (a consequence of Proposition 
\ref{prop:lambdaNP} below).
To such a series $f$, we attach the function $\Lambda(f) : \R \cup \{+\infty\} \to \R 
\cup \{+\infty\}$ defined by:
\[
\Lambda(f)(v) = \begin{cases} \log \left( \sup_{h \in B^-_E(e^v)} \Vert f(h) \Vert \right) & \mbox{if $f$ is defined on} B^-_E(e^v), \\
+\infty & \mbox{otherwise.} \end{cases}
\]
The following lemma is easy and left to the reader.

\begin{lem}
\label{lem:opLambda}
Let $f = \sum_{n \geq 0} f_n$ and $g = \sum_{n \geq 0} g_n$ be two 
series as above. Then
\begin{align*}
\Lambda(f+g) &\leq \max (\Lambda(f),\Lambda(g)), \\
\Lambda(f \times g) &\leq \Lambda(f)+\Lambda(g), \\
\Lambda(f \circ g) &\leq \Lambda(f) \circ \Lambda(g).
\end{align*}
\end{lem}

@@

Recall from \S \ref{ssec:stepbystep} that a SOMOS 4 sequence is a 
four-term inductive sequence defined by $u_n = \frac{u_{n-3} u_{n-1} 
+ u_{n-2}^2}{u_{n-4}}$ exhibiting the Laurent phenomenon.
We will focus on SOMOS sequences with values in $\Q_p$, and 
assume for simplicity that $u_0, u_1, u_2, u_3 \in \Z_p^\times$.
By the Laurent phenomenon, $u_n \in \Z_p$ for all $n$, and if $u_0, u_1, u_2, u_3$ 
are known with finite precision $O(p^N)$ then all $u_n$ are 
known with the same absolute precision. Algorithm \ref{algo:SOMOS} presented on 
page \pageref{algo:SOMOS} performs this computation.

@@

Recall that any 
ideal $I \subset K[\XX]$ admits a unique reduced Gr\"obner basis. We may
consider the map sending a family $(f_1, \ldots, f_s)$ of 
\emph{homogeneous}\footnote{This restriction will be convenient for us 
mainly because we are using the article \cite{vaccon:14a}, which deals only 
with homogeneous polynomials. It is however probably not essential.} 
polynomials of fixed degrees to the reduced Gr\"obner basis $(g_1, 
\ldots, g_t)$. It follows from \cite{vaccon:14a}*{Thm. 1.1} that this 
map is continuous at $(f_1, \ldots, f_s)$ provided that:
\begin{itemize}
\item the sequence $(f_1, \ldots, f_s)$ is regular
\item for all $i \in \{1, \ldots, s\}$, the ideal generated by
$f_1, \ldots, f_i$ is weakly-$w$ where $w$ denotes the fixed monomial
order (\emph{cf} \cite{vaccon:14a} for more detail).
\end{itemize}
A similar argument proves that it is actually differentiable at such
points. We are now going to compute the differential. For this we
remark that since the families $(f_1, \ldots, f_s)$ and $(g_1, \ldots,
g_t)$ generate the same ideal, we have a relation:
\begin{equation}
\label{eq:prodgrobner}
(g_1, \ldots, g_t) = (f_1, \ldots, f_s) \cdot A
\end{equation}
where $A$ is an $(s \times t)$ matrix with coefficients in $K[\XX]$.
Following Vaccon's construction, we see that the entries of $A$ 
can be chosen in such a way that they define differentiable functions.
Differentiating \eqref{eq:prodgrobner}, we get
$$(d g_1, \ldots, d g_t) =
(d f_1, \ldots, d f_s) \cdot A + (f_1, \ldots, f_s) \cdot dA.$$
Finally, since $(g_1, \ldots, g_t)$ is a \emph{reduced} Gr\"obner
basis, the above equality implies that $d g_i$ is the remainder in
the division of the $i$-th entry of the product
$(d f_1, \ldots, d f_s) \cdot A$ by the family $(g_1, \ldots, g_t)$.
